---
title: "dvp data processing file"
author: "Rayyan"
output: html_document
---

# Coding from DEP to clean the datasets for the DVP

```{r message = FALSE, warning = FALSE}

# Loading Libraries

library(tidyverse)
library(janitor)
library(readxl)
library(networkD3)
library(htmlwidgets)
library(highcharter)
library(openxlsx)
library(gplots)
library(gridExtra)
library(lmtest)
library(treemap)
library(leaflet)
library(gganimate)
library(av)
library(purrr)
library(broom)
library(kableExtra)

```

## Employee Earnings

```{r message = FALSE, warning = FALSE}
# Loading Datasets

employee_earnings <- 
  read_excel("data/63370_Table05.xlsx",
             sheet = "Data 5",
             skip = 5)

employee_earnings[1, 1] <- "Survey month"
employee_earnings[1, 2] <- "State and territory"
employee_earnings[1, 3] <- "Leave entitlements"
employee_earnings[1, 4] <- "Level of highest non-school qualification"
employee_earnings[1, 5] <- "Classification"
employee_earnings[1, 6] <- "Category"

employee_earnings[1, 7] <- "Persons_Median weekly_earnings"
employee_earnings[1, 8] <- "Persons_Median weekly_earnings_rse"
employee_earnings[1, 9] <- "Persons_Median hourly_earnings"
employee_earnings[1, 10] <- "Persons_Median hourly_earnings_rse"
employee_earnings[1, 11] <- "Persons_Employees_1000s"
employee_earnings[1, 12] <- "Persons_Employees_1000s_rse"

employee_earnings[1, 13] <- "Male_Median weekly_earnings"
employee_earnings[1, 14] <- "Male_Median weekly_earnings_rse"
employee_earnings[1, 15] <- "Male_Median hourly_earnings"
employee_earnings[1, 16] <- "Male_Median hourly_earnings_rse"
employee_earnings[1, 17] <- "Male_Employees_1000s"
employee_earnings[1, 18] <- "Male_Employees_1000s_rse"

employee_earnings[1, 19] <- "Female_Median weekly_earnings"
employee_earnings[1, 20] <- "Female_Median weekly_earnings_rse"
employee_earnings[1, 21] <- "Female_Median hourly_earnings"
employee_earnings[1, 22] <- "Female_Median hourly_earnings_rse"
employee_earnings[1, 23] <- "Female_Employees_1000s"
employee_earnings[1, 24] <- "Female_Employees_1000s_rse"


employee_earnings_tidy <- employee_earnings[-2,]

# Setting Column 1 observations as variable names
colnames(employee_earnings_tidy) <- employee_earnings_tidy[1, ]

# Removing original column 1
employee_earnings_tidy <- employee_earnings_tidy[-1, ]

employee_earnings_tidy[, 7:24] <- apply(employee_earnings_tidy[, 7:24], 2, function(x) as.numeric(as.character(x)))

```


```{r message = FALSE, warning = FALSE}

employee_earnings_tidy <- clean_names(employee_earnings_tidy)

employee_earnings_tidy[, c(7, 9, 13, 15, 19, 21)] <- 
  round(employee_earnings_tidy[, c(7, 9, 13, 15, 19, 21)], 1)

employee_earnings_tidy[, c(8, 10, 12, 14, 16, 18, 20, 22, 24)] <- 
  round(employee_earnings_tidy[, c(8, 10, 12, 14, 16, 18, 20, 22, 24)], 2)


```

```{r}

lm_data <- employee_earnings_tidy %>%
  filter(category == "Full-time")

# Fit a linear regression model
lm_model <- lm(persons_median_weekly_earnings ~ level_of_highest_non_school_qualification, data = lm_data)

# Check the summary of the regression model
summary(lm_model)

# Extract coefficients and p-values
coefficients <- coef(lm_model)
p_values <- coef(summary(lm_model))[, "Pr(>|t|)"]

#Print coefficients and p-values
print(coefficients)
print(p_values)

```


## Weekly Earning by Degree

```{r message = FALSE, warning = FALSE}

weekly_earning_degree <-
  read_csv("data/Median weekly earnings in main job, by highest educational qualification.csv",
           skip = 1)

weekly_earning_degree <- weekly_earning_degree %>%
  rename(Degree = '...1') %>%
  clean_names() %>%
  rename("2018" = x2018,
         "2023" = x2023)

weekly_earning_degree <- as.data.frame(weekly_earning_degree)

```

```{r}

degree_pivot <- weekly_earning_degree %>%
  pivot_longer(cols = -degree,
               names_to = "year",
               values_to = "weekly_earning")

```


## Weekly Earning by Gender

```{r message = FALSE, warning = FALSE}

weekly_earning_gender <- 
  read_csv("data/Median weekly earnings in main job, by sex and full-time or part-time status.csv",
           skip = 1)

weekly_earning_gender <- weekly_earning_gender %>%
  rename(Year = '...1') %>%
  clean_names()

weekly_earning_gender[, 2:5] <- 
  round(weekly_earning_gender[, 2:5], 1)

gender_filter <- weekly_earning_gender %>%
  filter(year >= 2018) %>%
  pivot_longer(cols = -year,
               names_to = "type",
               values_to = "amount")

summary_data <- gender_filter %>%
  group_by(type, year) %>%
  summarize(mean_amount = mean(amount))

# Plot the summarized data
# below

```

## Participation Rate and Unemployment Rate
```{r message = FALSE, warning = FALSE}

participation_rate <- 
  read_csv("data/Participation rate, Australia, August 1966 to January 2024, Seasonally adjusted.csv", skip =1)

participation_rate <- participation_rate %>%
  rename(Year = '...1') %>%
  clean_names()

```

```{r message = FALSE, warning = FALSE}

unemployment_rate <- 
  read_csv("data/Unemployment rate, Australia, August 1966 to January 2024, Seasonally adjusted.csv", skip = 1)

unemployment_rate <- unemployment_rate %>%
  rename(Year = '...1') %>%
  clean_names()
```


```{r}

workforce <- left_join(unemployment_rate,
                       participation_rate,
                       by = "year")

workforce <- workforce %>%
  rename(unemployment_persons = persons_percent.x) %>%
  rename(unemployment_male = males_percent.x) %>%
  rename(unemployment_female = females_percent.x) %>%
  rename(participation_persons = persons_percent.y) %>%
  rename(participation_male = males_percent.y) %>%
  rename(participation_female = females_percent.y)

workforce <- separate(workforce, year, into = c("month", "year"), sep = "-")

workforce <- workforce %>%
  mutate(year = as.numeric(year))

workforce$year <- ifelse(as.numeric(workforce$year) <= 24, 
                         as.character(as.numeric(workforce$year) + 2000), 
                         paste0("19", workforce$year))

workforce <- workforce %>%
  mutate(year = as.numeric(year))

workforce2 <- workforce %>%
  filter(year >= 2000)

```

```{r}
workforce_plot <- workforce %>%
  filter(year >= 2018 & year < 2024) %>%
  pivot_longer(cols = -c(month, year),
               names_to = "type",
               values_to = "percentage")

workforce_plot <- workforce_plot %>%
  group_by(type, year) %>%
  mutate(avg_value = mean(percentage))
        
```

```{r}

# Convert month to numeric representation
workforce2$month_numeric <- match(tolower(workforce2$month), tolower(month.abb))

# Convert month and year to a single date column
workforce2$date <- as.Date(ISOdate(workforce2$year, workforce2$month_numeric, 1))

# Remove month and year columns if you no longer need them
workforce2 <- subset(workforce2, select = -c(month, year, month_numeric))

# Reshape the data into a matrix format
heatmap_data <- as.matrix(workforce2[, c("unemployment_persons", "unemployment_male", "unemployment_female", "participation_persons", "participation_male", "participation_female")])

# Create the heatmap
# below

```

## Overall Migration

```{r message = FALSE, warning = FALSE}

overall_migration <- 
  read_excel("data/Overall Migration.xlsx",
           sheet = "Data1")

overall_migration <- overall_migration

overall_migration$Year <- format(overall_migration$Year, "%b-%Y")

overall_migration <- separate(overall_migration, Year, into = c("Month", "Year"), sep = "-")

```


```{r}

overall_migration_long <- overall_migration %>%
  filter(Year >= 2015) %>%
  pivot_longer(cols = -c(Month, Year), 
               names_to = c("Measurement Type", "State/Territory"), 
               names_pattern = "([^;]+) ;  (.+) ;", 
               values_to = "Value")

```


```{r}

# Filter data for the years 2015-2023
filtered_data <- overall_migration_long %>%
  filter(Year >= 2018) %>%
  filter(`State/Territory`!= "Australia") %>%
  rename(State = `State/Territory`)

# Calculate the total net overseas migration for each state
migration_totals <- aggregate(Value ~ State + `Measurement Type` + Year, 
                              data = filtered_data, 
                              sum)

migration_totals <- migration_totals %>%
  mutate(net_increase = ifelse(Value > 0, "Increase", "Decrease"))

```

# Data for tooltip
### Unemployment rate by state

```{r}

# Loading labour force data
tooltip_data_unemp <- read_xlsx("data/6202012.xlsx",
                                sheet = "Data2")

tooltip_unemp_states <- tooltip_data_unemp[, 151:175]
tooltip_unemp_states <- tooltip_unemp_states[-c(1:9), c(6,9,12,15,18,21,23,25)]
tooltip_unemp_states <- tooltip_unemp_states %>%
  clean_names()

tooltip_unemp_states <- tooltip_unemp_states %>%
  rename("New South Wales" = unemployment_rate_persons_new_south_wales_156) %>%
  rename("Victoria" = unemployment_rate_persons_victoria_159) %>%
  rename("Queensland" = unemployment_rate_persons_queensland_162) %>%
  rename("South Australia" = unemployment_rate_persons_south_australia_165) %>%
  rename("Western Australia" = unemployment_rate_persons_western_australia_168) %>%
  rename("Tasmania" = unemployment_rate_persons_tasmania_171) %>%
  rename("Northern Territory" = unemployment_rate_persons_northern_territory_173) %>%
  rename("Australian Capital Territory" = unemployment_rate_persons_australian_capital_territory_175)

# Adding date
start_date <- as.Date("1978-02-01")
end_date <- as.Date("2024-04-01")
date_sequence <- seq(from = start_date, to = end_date, by = "month")
tooltip_unemp_states$date <- date_sequence

tooltip_unemp_states <- tooltip_unemp_states %>%
  filter(date >= as.Date("2023-01-01"))

tooltip_unemp_states <- tooltip_unemp_states[, c(9, 1:8)]

# Converting to numeric
tooltip_unemp_states[,-1] <- lapply(tooltip_unemp_states[,-1], as.numeric)

# Converting to long format
state_averages <- tooltip_unemp_states %>%
  pivot_longer(cols = -date, 
               names_to = "state", 
               values_to = "unemployment")

# Getting mean
tp_final_1 <- state_averages %>%
  group_by(state) %>%
  summarise(average_unemployment = mean(unemployment))

```

### Participation rate by state


```{r}

# Same as above code chunk only for different statistic
tooltip_data_participation <- read_xlsx("data/6202012.xlsx",
                                sheet = "Data3")

tooltip_part_states <- tooltip_data_participation[, 120:141]
tooltip_part_states <- tooltip_part_states[-c(1:9), c(3,6,9,12,15,18,20,22)]
tooltip_part_states <- tooltip_part_states %>%
  clean_names()

tooltip_part_states <- tooltip_part_states %>%
  rename("New South Wales" = participation_rate_persons_new_south_wales_122) %>%
  rename("Victoria" = participation_rate_persons_victoria_125) %>%
  rename("Queensland" = participation_rate_persons_queensland_128) %>%
  rename("South Australia" = participation_rate_persons_south_australia_131) %>%
  rename("Western Australia" = participation_rate_persons_western_australia_134) %>%
  rename("Tasmania" = participation_rate_persons_tasmania_137) %>%
  rename("Northern Territory" = participation_rate_persons_northern_territory_139) %>%
  rename("Australian Capital Territory" = participation_rate_persons_australian_capital_territory_141)

start_date <- as.Date("1978-02-01")
end_date <- as.Date("2024-04-01")
date_sequence <- seq(from = start_date, to = end_date, by = "month")
tooltip_part_states$date <- date_sequence

tooltip_part_states <- tooltip_part_states %>%
  filter(date >= as.Date("2023-01-01"))

tooltip_part_states <- tooltip_part_states[, c(9, 1:8)]

tooltip_part_states[,-1] <- lapply(tooltip_part_states[,-1], as.numeric)

state_averages_part <- tooltip_part_states %>%
  pivot_longer(cols = -date, 
               names_to = "state", 
               values_to = "participation")

tp_final_2 <- state_averages_part %>%
  group_by(state) %>%
  summarise(average_participation = mean(participation))

```

### Migration Rates

```{r}

migration_stats <- overall_migration_long %>%
  group_by(`State/Territory`, `Measurement Type`) %>%
  summarise(Average_Value = mean(Value))

tp_final_3 <- migration_stats %>%
  pivot_wider(names_from = `Measurement Type`, values_from = Average_Value)

tp_final_3 <- tp_final_3[-1,]

tp_final_3 <- tp_final_3 %>%
  rename("state" = `State/Territory`)

```

### Average income

```{r}

tooltip_avg_incom <- employee_earnings_tidy %>%
  filter(classification == "Full-time or part-time status",
         leave_entitlements == "Total employees",
         level_of_highest_non_school_qualification == "Total")

tooltip_avg_incom <- tooltip_avg_incom[, c(2,6,7)]

tp_final_4 <- tooltip_avg_incom %>%
  group_by(state_and_territory, category) %>%
  summarise(average_value = mean(persons_median_weekly_earnings))

tp_final_4 <- tp_final_4 %>%
  pivot_wider(names_from = category,
              values_from = average_value)

tp_final_4 <- tp_final_4 %>%
  filter(state_and_territory != "Australia") %>%
  rename("state" = state_and_territory)

```

### Coordinates

```{r}

# Adding coordinates for the biggest cities in all of the states of Australia
# this is done for leaflet map purposes.
tp_final_5 <- data.frame(
  state = c("New South Wales", "Victoria", "Queensland", "South Australia", "Western Australia", "Tasmania", "Northern Territory", "Australian Capital Territory"),
  Latitude = c(-33.8688, -37.8136, -27.4698, -34.9285, -31.9505, -42.8821, -12.4634, -35.2809),
  Longitude = c(151.2093, 144.9631, 153.0251, 138.6007, 115.8605, 147.3272, 130.8456, 149.1300)
)

```



### Creating one dataset for tooltip data

```{r}

final_tooltip_data <- right_join(tp_final_1, 
                                 tp_final_2,
                                 by = "state")

final_tooltip_data <- right_join(final_tooltip_data, 
                                 tp_final_3,
                                 by = "state")

final_tooltip_data <- right_join(final_tooltip_data, 
                                 tp_final_4,
                                 by = "state")

final_tooltip_data <- right_join(final_tooltip_data, 
                                 tp_final_5,
                                 by = "state")

final_tooltip_data <- final_tooltip_data %>%
  clean_names()

final_tooltip_data[, 2:3] <- lapply(final_tooltip_data[, 2:3], round, digits = 2)
final_tooltip_data[, 4:7] <- lapply(final_tooltip_data[, 4:7], round, digits = 0)
final_tooltip_data[, 8:9] <- lapply(final_tooltip_data[, 8:9], round, digits = 1)

write.xlsx(final_tooltip_data, "data_clean/tooltip_data.xlsx")

```

# Data for Slide 2

```{r}

# Bar graph data for weekly earnings by employment type and by gender

slide2_data1 <- employee_earnings_tidy %>%
  filter(classification == "Full-time or part-time status",
         leave_entitlements == "Total employees",
         level_of_highest_non_school_qualification == "Total")

slide2_data1 <- slide2_data1[, c(2,6,13,19)]

data1_years <- rep(2014:2023, each = 18)

slide2_data1$year <- data1_years

slide2_data1 <- slide2_data1[, c(5, 1:4)]

slide2_data1 <- slide2_data1 %>%
  rename("male" = male_median_weekly_earnings) %>%
  rename("female" = female_median_weekly_earnings)

slide2_data1 <- slide2_data1 %>%
  pivot_longer(cols = -c(year, category, state_and_territory),
               names_to = "gender",
               values_to = "weekly_income")


write.xlsx(slide2_data1, "data_clean/weekly_earning_state.xlsx")

```


```{r}

# Data for violin plot
slide2_data2 <- read_xlsx("data/6202012.xlsx",
                                sheet = "Data3")

slide2_data2_part1 <- slide2_data2[, 117:141]
slide2_data2_part1 <- slide2_data2_part1[-c(1:9), c(3,6,9,12,15,18,21,23,25)]
slide2_data2_part1 <- slide2_data2_part1 %>%
  clean_names()

slide2_data2_part1 <- slide2_data2_part1 %>%
  rename("Australia" = participation_rate_persons_australia_119) %>%
  rename("New South Wales" = participation_rate_persons_new_south_wales_122) %>%
  rename("Victoria" = participation_rate_persons_victoria_125) %>%
  rename("Queensland" = participation_rate_persons_queensland_128) %>%
  rename("South Australia" = participation_rate_persons_south_australia_131) %>%
  rename("Western Australia" = participation_rate_persons_western_australia_134) %>%
  rename("Tasmania" = participation_rate_persons_tasmania_137) %>%
  rename("Northern Territory" = participation_rate_persons_northern_territory_139) %>%
  rename("Australian Capital Territory" = participation_rate_persons_australian_capital_territory_141)

start_date1 <- as.Date("1978-02-01")
end_date1 <- as.Date("2024-04-01")
date_sequence1 <- seq(from = start_date1, to = end_date1, by = "month")
slide2_data2_part1$date <- date_sequence1

slide2_data2_part1 <- slide2_data2_part1 %>%
  filter(date >= as.Date("2014-01-01"))

slide2_data2_part1 <- slide2_data2_part1[, c(10, 1:9)]

slide2_data2_part1[,-1] <- lapply(slide2_data2_part1[,-1], as.numeric)

slide2_data2_part1 <- slide2_data2_part1 %>%
  mutate(year = format(date, "%Y")) %>%
  group_by(year) %>%
  summarize(across(c(Australia, `New South Wales`, Victoria, Queensland, 
                     `South Australia`, `Western Australia`, Tasmania, `Northern Territory`, 
                     `Australian Capital Territory`), 
                   mean, 
                   na.rm = TRUE)) %>%
  ungroup()

slide2_data2_final <- slide2_data2_part1 %>%
  pivot_longer(cols = -year, 
               names_to = "state", 
               values_to = "participation")

write.xlsx(slide2_data2_final, "data_clean/overall_participation_rate.xlsx")

```

# Downloading clean datasets for smoother shiny app development
# Testing visualizations and models

```{r}

# Downloading migration total dataframe

write_csv(migration_totals, "data_clean/migration_totals.csv")

```

```{r}

degree_pivot %>%
ggplot(aes(x = year, y = weekly_earning, fill = degree, group = degree)) + 
  geom_area(alpha = 0.6, 
            size = 1, 
            colour = "black", 
            stat = 'identity') + 
  theme_minimal() +
  labs(title = "Area Plot of Weekly Earnings by Degree and Year", 
       x = "Year", 
       y = "Weekly Earning") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_fill_brewer(palette = "Set1")

```

```{r}

table(employee_earnings_tidy$level_of_highest_non_school_qualification)


slide3_data1 <- employee_earnings_tidy %>%
  filter(classification == "Full-time or part-time status",
         leave_entitlements == "Total employees")

slide3_data1 <- slide3_data1[, c(2,4,6,13,19)]

slide3_data1_years <- rep(2014:2023, each = 180)

slide3_data1$year <- slide3_data1_years

slide3_data1 <- slide3_data1[, c(6, 1:5)]

slide3_data1 <- slide3_data1 %>%
  rename("male" = male_median_weekly_earnings) %>%
  rename("female" = female_median_weekly_earnings)

slide3_data1 <- slide3_data1 %>%
  pivot_longer(cols = -c(year, 
                         category,
                         state_and_territory, 
                         level_of_highest_non_school_qualification),
               names_to = "gender",
               values_to = "weekly_income")

states <- unique(slide3_data1$state_and_territory)

state_data_list <- split(slide3_data1, slide3_data1$state_and_territory)

write.xlsx(slide3_data1, "data_clean/state_data.xlsx")

```


```{r}

lm_data1 <- slide3_data1 %>%
  filter(category == "Full-time")

# Step 2: Split the data by state
state_data_list <- split(lm_data1, lm_data1$state_and_territory)

# Step 3: Function to fit the linear model and extract coefficients and p-values
fit_lm_extract <- function(data) {
  lm_model <- lm(weekly_income ~ level_of_highest_non_school_qualification, data = data)
  tidy_model <- tidy(lm_model)
  
  # Select and rename relevant columns
  tidy_model %>%
    select(term, estimate, p.value) %>%
    rename(
      education_type = term,
      coefficient = estimate,
      pvalue = p.value
    )
}

# Step 4: Apply the function to each state and add a column for state name
results <- map_dfr(names(state_data_list), function(state) {
  state_data <- state_data_list[[state]]
  state_results <- fit_lm_extract(state_data)
  state_results %>%
    mutate(state = state)
})

# Step 5: Print the final combined results

results <- results %>%
  mutate(education_type = str_trim(sub("level_of_highest_non_school_qualification", "", education_type)))

print(results)

# Download

write.xlsx(results, "data_clean/lm_model_full_time.xlsx")

```

```{r}

lm_part_time <- read.xlsx("data_clean/lm_model_part_time.xlsx")
lm_full_time <- read.xlsx("data_clean/lm_model_full_time.xlsx")


head(lm_full_time)
head(lm_part_time)

lm_full_time %>%
  filter(state == "Victoria") %>%
  kable(caption = paste("Linear Regression Model for Victoria")) %>%
  row_spec(0, background = "white")



```

```{r}

state_data_list$Victoria %>%
  ggplot(aes(x = year, y = weekly_income, fill = level_of_highest_non_school_qualification)) +
  geom_violin(trim = FALSE, scale = "width") +
  theme_bw() +
  scale_x_continuous(breaks = seq(2014, 2023, by = 1)) +
  scale_y_continuous(breaks = seq(
    min(slide3_data1$weekly_income),
    max(slide3_data1$weekly_income), 
    by = 500)) +
  labs(title = paste("Violin Plot of Income Distribution by Education Type in", 
                     state_data_list$Victoria$state_and_territory),
       x = "Year",
       y = "Weekly Income",
       fill = "Education Type")

chunk38 <- state_data_list[["Victoria"]]

table(chunk38$state_and_territory)

```



